---
title: "october 18 final project ver"
author: "MJ" "Liv.Thrower"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#installing a Grid Extra pacakge to make formatting our plots easier - this allows us to show plots side by side 
#install.packages("gridExtra")
library(gridExtra)
```

loading tidyverse, uploading the pollution download data
```{r}
library(tidyverse)
library(dplyr)
final.proj <- read.csv(file.choose())
view(final.proj)
```
#Country : Name of the country
#City : Name of the city
#AQI Value : Overall AQI value of the city
#AQI Category : Overall AQI category of the city
#CO AQI Value : AQI value of Carbon Monoxide of the city
#CO AQI Category : AQI category of Carbon Monoxide of the city
#Ozone AQI Value : AQI value of Ozone of the city
#Ozone AQI Category : AQI category of Ozone of the city
#NO2 AQI Value : AQI value of Nitrogen Dioxide of the city
#NO2 AQI Category : AQI category of Nitrogen Dioxide of the city
#PM2.5 AQI Value : AQI value of Particulate Matter with a diameter of 2.5 micrometers or less of the city
#PM2.5 AQI Category : AQI category of Particulate Matter with a diameter of 2.5 micrometers or less of the city





classing the data 
```{r}
class(final.proj$Country)               #character
class(final.proj$City)                  #character
class(final.proj$AQI.Value)             #integer
class(final.proj$AQI.Category)          #character
class(final.proj$CO.AQI.Value)          #integer
class(final.proj$Ozone.AQI.Value)       #integer
class(final.proj$Ozone.AQI.Category)    #character
class(final.proj$NO2.AQI.Value)         #integer
class(final.proj$NO2.AQI.Category)      #character
class(final.proj$PM2.5.AQI.Value)       #integer
class(final.proj$PM2.5.AQI.Category)    #character
```

Making an object for all the data in the usa

```{r}
us.data <- final.proj %>%
  filter(Country == "United States of America") %>%
  na.omit() #just in case
```

```{r}
#Top 5 highest cities of AQI Value in USA.
us.data_test <- final.proj %>%
  select(Country,City, AQI.Value) %>%
    filter(Country == "United States of America") %>%
  arrange(desc(AQI.Value)) %>%
slice(1:15)

```




Overall AQI Values 

Including outlier (Durango, COLORADO @ AQI = 500)


makign an object to contain this info (for side-by-side comparison ease later on)
```{r}
us.cities.AQI <- ggplot(data = us.data, aes(x = City, y = AQI.Value, color = AQI.Value)) + 
  geom_point() +
  scale_y_continuous(
    breaks = seq(0, 500, by = 50),
    limits = c(0, 500)
  ) +
  scale_color_gradient(low = "lightgreen", high = "brown") +
  labs(title = "AQI Values for Various US Cities", x = "Cities in the United States", y = "AQI Value") +
  theme(axis.text.x = element_blank()) #keeps from having all the city names listed (too many to create a visible/accessible x-axis)

print(us.cities.AQI)
```


Omitting the AQI=500 outlier point - now it is just AQI values of different US cities
```{r}
us.cities.AQI.noOutlier <- ggplot(data = us.data, aes(x = City, y = AQI.Value, color = AQI.Value)) + 
  geom_point() +
  scale_y_continuous(             #formatting the y-axis
    breaks = seq(0, 175, by = 25),
    limits = c(0, 175)        
  ) + #the following makes a color gradient 
  scale_color_gradient(low = "lightgreen", high = "darkgreen", breaks = seq(0, 175, by = 25), limits = c(0, 175)) +
  labs(title = "AQI Values for Various US Cities*", x = "Cities in the United States", y = "AQI Value") +
  theme(axis.text.x = element_blank()) #keeps from having all the city names listed (too many to create a visible/accessible x-axis)
print(us.cities.AQI.noOutlier)

#the warning saying there is a missing value is due to the exclusion of the outlier while using the original us city data set
```
if we want to see these plots side by side:
```{r}
grid.arrange(us.cities.AQI, us.cities.AQI.noOutlier, ncol = 2)

```


defining custom colors for use in plots
```{r}
custom_colors <- c(
  "Good" = "green",
  "Moderate" = "yellow",
  "Unhealthy for Sensitive Groups" = "orange",
  "Unhealthy" = "red",
  "Very Unhealthy" = "purple",
  "Hazardous" = "brown"
)

```


```{r}
# Top 15 plotted aqi in USA

top_aqi_usa <- ggplot(data = us.data_test, aes(x = City, y = AQI.Value )) +
  coord_flip() +
 geom_bar( stat = 'identity', fill = 'seagreen3') 

top_aqi_usa



```


OZONE PLOTS


```{r}
us.cities.ozone.scatter <- ggplot(data = us.data, aes(x = City, y = Ozone.AQI.Value, color = Ozone.AQI.Value)) + 
  geom_point() +
  scale_y_continuous(             #formatting the y-axis
    breaks = seq(0, 150, by = 25),
    limits = c(0, 150)        
  ) + #the following makes a color gradient 
  scale_color_gradient(low = "green", high = "brown", breaks = seq(0, 150, by = 25), limits = c(0, 150)) +
  labs(title = "Ozone AQI Values for Various US Cities", x = "Cities in the United States", y = "Ozone AQI Value") +
  scale_fill_manual(values = custom_colors) + theme(
        plot.title = element_text(size = 10))  

print(us.cities.ozone.scatter)
```

summarizing data!!
```{r}
us.ozone.cat <- us.data %>%
  group_by(Ozone.AQI.Category) %>%
  summarize(Count = n(), .groups = "drop")
```


bar plot for ozone

```{r}
us.cities.ozone.bar <- ggplot(us.ozone.cat, aes(x = Ozone.AQI.Category, y = Count, fill = Ozone.AQI.Category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Ozone AQI Categorical Rating for Various US Cities",
    x = "AQI Categories",
    y = "Count"
  ) +   scale_fill_manual(values = custom_colors) + theme(
        plot.title = element_text(size = 10)) 
print(us.cities.ozone.bar)
```

this allows us to see, side by side, the Ozone Values + Categorical Ranking for the US

```{r}
grid.arrange(us.cities.ozone.scatter, us.cities.ozone.bar, ncol = 2)
```


NO2



no2 scatter plot

```{r}
us.cities.no2.scatter <- ggplot(data = us.data, aes(x = City, y = NO2.AQI.Value, color = NO2.AQI.Value)) + 
  geom_point() +
  scale_y_continuous(             #formatting the y-axis
    breaks = seq(0, 150, by = 25),
    limits = c(0, 150)        
  ) + #the following makes a color gradient 
  scale_color_gradient(low = "green", high = "brown", breaks = seq(0, 150, by = 25), limits = c(0, 150)) +
  labs(title = "NO2 AQI Values for Various US Cities", x = "Cities in the United States", y = "NO2 AQI Value") +
   scale_fill_manual(values = custom_colors) +   theme(axis.text.x = element_blank(),
        plot.title = element_text(size = 10))
print(us.cities.no2.scatter)
```

summarizing data for no2, sorting by group, and counting (for use in bar plot)

```{r}
us.no2.cat <- us.data %>%
  group_by(NO2.AQI.Category) %>%
  summarize(Count = n(), .groups = "drop")
```


bar plot for no2


```{r}
us.cities.no2.bar <- ggplot(us.no2.cat, aes(x = NO2.AQI.Category, y = Count, fill = NO2.AQI.Category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "NO2 AQI Categorical Rating for Various Cities",
    x = "AQI Categories",
    y = "Count"
  ) +   scale_fill_manual(values = custom_colors) +
 scale_fill_manual(values = custom_colors) + theme(
        plot.title = element_text(size = 10)) 
print(us.cities.no2.bar)
```


this allows us to see, side by side, the no2 Values + Categorical Ranking for the US

```{r}
grid.arrange(us.cities.no2.scatter, us.cities.no2.bar, ncol = 2)
```

  PM2.5 DATA
  
  
making an object for PM2 scatter plot to be stored in <3


note: there is again an outlier, where PM2.5 = 500 (Durango, again)

plot for PM2.5 w/outlier
```{r}
us.cities.pm2.5.scatter <- ggplot(data = us.data, aes(x = City, y = PM2.5.AQI.Value, color = PM2.5.AQI.Value)) + 
  geom_point() +
  scale_y_continuous(             #formatting the y-axis
    breaks = seq(0, 500, by = 50),
    limits = c(0, 500)        
  ) + #the following makes a color gradient 
  scale_color_gradient(low = "green", high = "brown", breaks = seq(0, 500, by = 50), limits = c(0, 500)) +
  labs(title = "PM2.5 AQI Values for Various US Cities", x = "Cities in the United States", y = "PM2.5 AQI Value") +
   scale_fill_manual(values = custom_colors) +   theme(axis.text.x = element_blank(),
        plot.title = element_text(size = 10))
print(us.cities.pm2.5.scatter)
```

Plot for us PM2.5 w/o outlier (Durango = 500)

```{r}
us.cities.pm2.5.noOutlier.scatter <- ggplot(data = us.data, aes(x = City, y = PM2.5.AQI.Value, color = PM2.5.AQI.Value)) + 
  geom_point() +
  scale_y_continuous(             #formatting the y-axis
    breaks = seq(0, 175, by = 25),
    limits = c(0, 175)        
  ) + #the following makes a color gradient 
  scale_color_gradient(low = "lightgreen", high = "darkgreen", breaks = seq(0, 175, by = 25), limits = c(0, 175)) +
  labs(title = "PM2.5 AQI Values for Various US Cities", x = "Cities in the United States", y = "PM2.5 AQI Value") +
   scale_fill_manual(values = custom_colors) +   theme(axis.text.x = element_blank(),
        plot.title = element_text(size = 10))
print(us.cities.pm2.5.noOutlier.scatter)

#the warning for 1 row containing missing values removed is the outlier being omitted from the plot for not fitting w/in the bounds. this is what we wanted.
```
if we want to see these plots side by side:
```{r}
grid.arrange(us.cities.pm2.5.scatter, us.cities.pm2.5.noOutlier.scatter, ncol = 2)
```

now: summarize the data for pm2/5
```{r}
us.pm2.5.cat <- us.data %>%
  group_by(PM2.5.AQI.Category) %>%
  summarize(Count = n(), .groups = "drop")
```


bar plot for pm2.5

```{r}
us.cities.pm2.5.bar <- ggplot(us.pm2.5.cat, aes(x = PM2.5.AQI.Category, y = Count, fill = PM2.5.AQI.Category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "PM2.5 AQI Categorical Rating for Various Cities",
    x = "AQI Categories",
    y = "Count"
  ) +   scale_fill_manual(values = custom_colors) +
 scale_fill_manual(values = custom_colors) + theme(
        plot.title = element_text(size = 10), axis.text.x = element_text(size = 6)) 
#adjusting the x-axis labels b/c the words are overlapping 
print(us.cities.pm2.5.bar)
```
display the PM2.5 scatter plots side by side:

```{r}
grid.arrange(us.cities.pm2.5.scatter, us.cities.pm2.5.bar, ncol = 2)
```


```{r}
grid.arrange(us.cities.AQI, us.cities.AQI.noOutlier, us.cities.ozone.scatter, us.cities.ozone.bar, us.cities.no2.scatter, us.cities.no2.bar, us.cities.pm2.5.noOutlier.scatter,us.cities.pm2.5.bar, ncol = 4)  

```
# it might be cool to use a smaller data set to develop statistical figure. This way we can allow the dots to become their own country - better for a reader to understand visually, and then we can make comparisons using the method facet wrap to make more conclusions. Lets focus on all the cities of the northeast, implement this change in the first R chunk, using filter and select commands. 
